---
title: "Automatic vs Manual Transmission on MPG"
author: "Xinchun Tian"
subtitle: Regression Models Course Project
output: pdf_document
---

# Executive Summary
High dimensional data analysis has become increasingly frequent and important in diverse fields of sciences, engineering, and humanities, ranging from genomics and health sciences to economics, finance and machine learning [1]. Generally the inclusion of multiple correlated variables in the model will
make the individual regression coefficients to be estimated unstably, i.e.
with big variance. If we only use a small subset of variables, then the
models are likely to be biased. Often we would like to achieve a balance
between the variance and the bias [2]. This project analyzes the Motor Trend Car Road Tests (mtcars) dataset, a collection of cars in exploring the relationship between a set of variables and miles per gallon (MPG) (outcome), We are trying to answer the following two questions:  
1. Is an automatic or manual transmission better for MPG?  
2. Quantify the MPG difference between automatic and manual transmissions.  
Two models are selected to demonstrate the importance of model selection, one is automatic/manual as the only predictor, and the other is the minimal adequte model.


```{r echo=FALSE}
suppressPackageStartupMessages(library(datasets))
suppressPackageStartupMessages(library(MASS))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(gridExtra))
data(mtcars)
```

# Exploratory Data Analysis
The data look clean and there are no missing values to impute (Appendix A). However we need to transform some of the variables to factor variables (cyl, vs, am, gear, carb). The table shows the first 5 observations of the mtcars dataset.

```{r echo=FALSE, fig.align="center"}
mtcars <- mutate(mtcars, cyl = as.factor(cyl), vs = as.factor(vs),
                 gear = as.factor(gear), carb = as.factor(carb), am = as.factor(am))
grid.table(mtcars[1:5,])
```

# Regression Analysis
```{r echo=FALSE, fig.align="center"}
library(ggplot2)
fit <- lm(mpg ~ am, mtcars)
```
If we naively took "am" (0 automatic, 1 manual) as the predictor and "mpg" as the response, we may mistakenly conclude that, on average, the manual transmission cars have 7.25 higher MPGs than automatic transmission cars as shown in the table below. 
However if you look at the "Residuals Fitted (1)" plot in Appendix B, there is a clear pattern. Also, the $R^2$=`r round(summary(fit)$r.squared,3)` is small, in other words, the model considering single predictor "am" only explains 36% of the variability of the response data around its mean. There are confounders that we need to include into the regression model.
```{r echo=FALSE, fig.align="center"}
grid.table(round(summary(fit)$coef,3))
```

## Model Selection
We begin with throwing all the predictors into the regression model and performs stepwise model selection by AIC (Appendix C).
The final model is using "cyl", "hp", "wt" and "am" as the predictors. The coefficients are shown in the following table. The "Residuals vs Fitted (2)" plot in Appendix B does not show any obvious patterns.
```{r echo=FALSE, fig.align="center"}
fit <- lm(mpg ~ cyl + hp + wt + am, data = mtcars)
grid.table(round(summary(fit)$coef,3))
confidence_interval <- summary(fit)$coef[6,1] + c(-1,1)*qt(0.975,df=summary(fit)$df[2])*summary(fit)$coef[6,2]
```

The slope of "am" is `r round(summary(fit)$coef[6,1], 3)`, which means, on average, holding other variables constant the manual transmission cars have `r round(summary(fit)$coef[6,1], 3)` higher MPGs than automatic transmission cars. However the "p-value" (`r round(summary(fit)$coef[6,4], 3)`) is large, so we failed to reject the null-hypotheis that the manual transmission cars have the same MPGs as that of automatic transmission cars. The 95% confidence interval is [`r round(confidence_interval,3)`] also includes 0.

# Summary
After analyzing the "mtcars" dataset, we conclude:  
1. There is no significant diffrence on MPG by automatic or manual transmission,  
2. The MPG difference between automatic and manual transmissions is [`r round(confidence_interval,3)`] at 95% C.L.  

# Reference
[1] Jianqing Fan and Jinchi Lv, Stat Sin. 2010 Jan; 20(1): 101–148.  
[2] http://www.ams.sunysb.edu/~xing/AMS578/Documents/Chp9.pdf

\newpage

# Apendeix
## Apendeix A
```{r echo = FALSE, fig.height=6, fig.width=8, fig.align="center"}
str(mtcars)
summary(mtcars)
par(mar=c(5, 4, 4, 1))
pairs(mtcars)
```

## Apendix B
```{r echo = FALSE}
par(mfrow=c(1,2))
plot(predict(lm(mpg ~ am, mtcars)), resid(lm(mpg ~ am, mtcars)), col=rgb(255,0,100,100,maxColorValue=255), pch=16,
     xlab="Fitted values", ylab="Residuals", main="Residuals vs Fitted (1)")
plot(predict(lm(mpg ~ cyl + hp + wt + am, mtcars)), resid(lm(mpg ~ cyl + hp + wt + am, mtcars)), col=rgb(0,255,100,100,maxColorValue=255), pch=16,
     xlab="Fitted values", ylab="Residuals", main="Residuals vs Fitted (2)")
```

## Appendix C
```{r echo = FALSE}
model <- lm(mpg ~ ., data=mtcars)
#summary(model)$coef
model_selection <- stepAIC(model, direction = 'both', trace = F)
model_anova <- model_selection$anova
model_anova
```