---
title: "Data Science Specification - Statistical Inference Homework"
output: html_document
---

- 1.1
P(A) = 0.1  
P(B) = 0.9  
P(A or B) = 0.15  
P(A and B) = 0.1 + 0.09 - 0.15

- 1.2 
0.50

- 1.3  
odds d = p/(1-p), so p = d/(1+d), so -X*d + Y*(1-p) = 0

- 1.4
```{r}
(-4-0)^2*0.2 + (1-0)^2*0.8
```

- 1.5  
If X and Y are independent, then
$$Var(\bar{X}-\bar{Y}) = Var(\bar{X}) + Var(\bar{Y})$$

- 1.6  
$E[\frac{X}{\sigma}]=\frac{E[X]}{\sigma}$,
$Var(\frac{X}{\sigma})=\frac{1}{\sigma^2}Var(X)$=1

- 1.8  
```{r}
p <- c(.1, .2, .3, .4)
x <- 2 : 5
mns <- sum(p*x)
sum(p*(x-mns)^2)
sum(x ^ 2 * p) - sum(x * p) ^ 2
```

- 2.1  
P(A) = 0.12  
P(A|B) = 0.90  
P(B) = 0.20  
P(A|B) = P(A & B) / P(B) <= P(A)P(B)/P(B) <= P(A) = 0.12

- 2.2  
```{r}
pnorm(93, mean=100, sd=10, lower.tail=T)
```

- 2.3  
$P(A) = \sum_n P(A \cap B_n) = \sum_n P(A|B_n)P(B_n)$
P(+|D) = 0.93  
P(-|~D) = 0.88  
P(D) = 0.05  
P(D|+) = P(+|D)P(D)/(P(+|D)P(D) + P(+|~D)P(~D))  
P(D|+) = P(+|D)P(D)/(P(+|D)P(D) + (1-P(-|~D))(1-P(D))) 
P(~D|-) = P(-|~D)P(~D)/(P(-|~D)P(~D) + P(-|D)P(D))
```{r}
0.93/(0.93+(1-0.88)*0.95)
0.88*0.95/(0.88*0.95+0.07*0.05)
```

- 2.4  
```{r}
qnorm(0.95, mean=100, sd=10, lower.tail = T)
```

- 2.5  
```{r}
qnorm(0.95, mean=100, sd=10/sqrt(50), lower.tail = T)
```

- 2.5  
```{r}
pbinom(4, size=6, prob=0.5, lower.tail = F)*100
```

- 2.6  
```{r}
# using the CLT, the sample distribution is gaussian.
pnorm(0.51, mean=0.5, sd=sqrt(1/12)/sqrt(100), lower.tail = F)
```

- 2.7  
```{r}
mns  <- (1+2+3+4+5+6)/6
vars <- (1^2+2^2+3^2+4^2+5^2+6^2)/6 - mns^2
vars/10

mean((1:6 - 3.5)^2/10)
```

- 2.8  
```{r}
ppois(20, 33, lower.tail = T)*100
```


- 3.1  
```{r}
data(mtcars)
n <- nrow(mtcars)
mean(mtcars$mpg) + c(-1,1)*qt(0.975,df=n-1)*sd(mtcars$mpg)/sqrt(n)

attach(mtcars)
t.test(mpg)

t.test(mpg)$conf.int
```

- 3.2  
```{r}
# why df = 9-1 instead of 9+9-2 because they are paired.
x <- qt(0.975,df=8)*1/sqrt(9)
x
```

- 3.3  
An independent group Student's T interval is used instead of a paired T interval when:  
A. The observations are paired between the groups.  
B. The observations between the groups are naturally assumed to be statistically independent (X)    
C. As long as you do it correctly, either is fine.  
D. More details are needed to answer this question  

- 3.4  
```{r}
data(mtcars)
mtcars4 <- mtcars[mtcars$cyl==4,]
mtcars6 <- mtcars[mtcars$cyl==6,]
n1 <- nrow(mtcars4)
n2 <- nrow(mtcars6)
m1 = mean(mtcars4$mpg)
m2 = mean(mtcars6$mpg)
var1 = var(mtcars4$mpg)
var2 = var(mtcars6$mpg)
sp = sqrt(((n1-1)*var1 + (n2-1)*var2)/(n1+n2-2))
m1 -m2 + c(-1,1)*qt(0.975,df=n1+n2-2)*sp*sqrt(1/n1+1/n2)
t.test(mtcars4$mpg, mtcars6$mpg, paired=F, var.equal=T)
```

- 3.5  
```{r}
n1 <- 9
n2 <- 9
m1 = 3
m2 = 1
sd1 = 1.5
sd2 = 1.8
sp = sqrt(((n1-1)*sd1^2 + (n2-1)*sd2^2)/(n1+n2-2))
sp^2
```
One point in this question is that the set of cars is a reasonable IID sample from a poplulation of cars we are interested in.library


- 3.6  
For Binomial data the maximum likelihood estimate for the probability of a success is  
A. The proportion of successes (X)  
B. The proportion of failures  
C. A shrunken version of the proportion of successes  
D. A shrunken version of the proportion of failures  


-4.1  
```{r}
x <- mean(mtcars$mpg)
sd <- sd(mtcars$mpg)
x + qnorm(0.975)*sd/sqrt(nrow(mtcars))
```
```{r}
mn <- mean(mtcars$mpg)
s <- sd(mtcars$mpg)
z <- qnorm(.05)
mu0 <- mn - z * s / sqrt(nrow(mtcars))
```

- 4.2  
```{r}
data(mtcars)
mtcars4 <- mtcars[mtcars$cyl==4,]
mtcars6 <- mtcars[mtcars$cyl==6,]
n1 <- nrow(mtcars4)
n2 <- nrow(mtcars6)
m1 = mean(mtcars4$mpg)
m2 = mean(mtcars6$mpg)
var1 = var(mtcars4$mpg)
var2 = var(mtcars6$mpg)
sp = sqrt(((n1-1)*var1 + (n2-1)*var2)/(n1+n2-2))
m1 - m2 + c(-1,1)*qt(0.975,df=n1+n2-2)*sp*sqrt(1/n1+1/n2)
t.test(mtcars4$mpg, mtcars6$mpg, paired=F, var.equal=F, alternative = "two.sided")
```

- 4.3  
```{r}
3.0 + c(-1,1)*qnorm(0.975)*1.1/sqrt(100)
```

- 4.4  
```{r}
pbinom(54, size=100, prob=0.5, lower.tail = F)
pnorm(.55, mean=0.5, sd = sqrt(0.5*(1-0.5)/100), lower.tail = F)
```

- 4.5   
```{r}
ppois(15800, 520*30, lower.tail = F)
# it should be 
ppois(15800-1, 520*30, lower.tail = F)
pnorm(15800/30, mean=520, sd=sqrt(520/30), lower.tail = F)
```

- 4.6  
```{r}
n1 = 100
n2 = 100
m1 = 10
m2 = 11
sd = 4
sdem <- sqrt(4^2)*sqrt(1/n1 + 1/n2)
z <- (m2 - m1)/sdem
2*pnorm(z, lower.tail = F)
```

- 4.7 
```{r}
pnorm(10+qnorm(0.95)*4/sqrt(100), mean=11, sd = 4/sqrt(100), lower.tail = F)
power.t.test(n = 100, delta = 1, sd=4, type = "one.sample",  alt = "one.sided")$power
```

- 4.8
```{r}
power.t.test(power=0.8, delta = 0.01, sd=0.04, type = "one.sample",  alt = "one.sided")$n
```

- 4.9  
```{r}
data(mtcars)
mtcars8 <- mtcars[mtcars$cyl==8,]
mtcars6 <- mtcars[mtcars$cyl==6,]
n1 <- nrow(mtcars8)
n2 <- nrow(mtcars6)
m1 = mean(mtcars8$mpg)
m2 = mean(mtcars6$mpg)
var1 = var(mtcars8$mpg)
var2 = var(mtcars6$mpg)
sp = sqrt(((n1-1)*var1 + (n2-1)*var2)/(n1+n2-2))
m1 - m2 + c(-1,1)*qt(0.975,df=n1+n2-2)*sp*sqrt(1/n1+1/n2)
t.test(mtcars8$mpg, mtcars6$mpg, paired=F, var.equal=T, alternative = "two.sided")
pnorm((m2-m1)/(sp*sqrt(1/n1+1/n2)),lower.tail = F)
sp
```